---
title: "Software dependancy analysis"
date: "11/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

```{r libraries, include=FALSE}
library(tidyverse)
library(here) # to read csv files with the function here
library(lubridate)
library(semver) # to extract the version numbers into separate columns
library(formattable) # to format tables
library(scales)
library(reshape)
```

## Context

Open source projects rely on libraries that get updated regularly. While not all updates require code changes, some do. 

According to [semantic versioning](https://semver.org), versions are numbered incrementally according to the type of change. A "MAJOR version" signifies incompatible API changes, a "MINOR version" signifies added functionality in a backwards compatible manner, and a "PATCH version" signifies backwards compatible bug fixes.

This project explores the number and frequency of updates in a project currently under development. It aims to shed light on the update requirements once the project will be in production.

## Methods

### Writing the data to a JSON file
The [dependency manifest file](https://github.com/canada-ca/tracker/blob/003a3d6e0452fb9a248a7d51fd5556bc03dfbcc6/frontend/package.json) from the frontend of Government of Canada's Tracker project was used as a representative example of a modern software project.

The manifest was retrieved using [curl](https://curl.se/) with the following command on November 12, 2020.
```
curl -sL "https://raw.githubusercontent.com/canada-ca/tracker/003a3d6e0452fb9a248a7d51fd5556bc03dfbcc6/frontend/package.json" > data/raw/tracker-frontend-package.json
```
### Converting the data to a CSV file
Using the following [Bash](https://www.gnu.org/software/bash/) command, package names (both development and production) were extracted from the `tracker-frontend-package.json` and details for each was retrieved from the npm API.

The command uses [jq](https://stedolan.github.io/jq) to select dependency name from the json and format the API response as CSV. Once again `curl` was used to retrieve data from the API.
```
for pkg in $(cat data/raw/tracker-frontend-package.json | jq '.dependencies * .devDependencies | keys[]' | tr -d \"); do $(curl -sL "https://registry.npmjs.org/$pkg" | jq -r --arg  name "$pkg" '.time | to_entries[] | [$name, .key, .value] | @csv' >> data/processed/nodejs.csv && sleep 1 ); done
```

The CSV contained the name of the package, the version and a time stamp.

### Cleaning the CSV file

The data needed a little attention before conducting analysis. The reader is referred to the [Rmarkdown file](https://github.com/caromimo/dependency-analysis/blob/master/index.Rmd) for code and details. 

Briefly, the data was cleaned in the four main steps:

1. read the CSV file and specify column names;
2. removed rows that did not contain data - for each package, the "time" returned a series of keys (version numbers but also a "created" and a "modified") and values (time stamp) - it was assumed that the row with "created" and the earliest version of a package were on the same day and that the "modified" and the latesst version of a package were on the same day; 
3. extracted the date from the time stamp, extracted the year as a factor, extracted the month as a factor, extracted the day of the week as a factor and parsed the version number to only include the first three numbers and the dots separating the numbers; and
4. extracted the version numbers into separate columns for major, minor and patch versions using the semver library.

```{r, include=FALSE}
packages <- read_csv(here("data", "processed", "nodejs.csv"), col_names = c("package_name", "version", "time_stamp"))
```

```{r, include=FALSE}
data <- packages %>%
  filter(version != "created") %>%
  filter(version != "modified")
```

```{r, include=FALSE}
data <- data %>%
  mutate(
    date = as.Date(time_stamp, "%Y/%m/%d"),
    year = as.factor(year(date)),
    month = as.factor(month(date)),
    weekday = as.factor(weekdays(date)),
    version = str_replace(version, "^(\\d{1,2}.\\d{1,2}.\\d{1,2})(\\w+)$", "\\1-\\2")
  )
```

```{r, include=FALSE}
data <- add_column(
  data,
  pull(
    data, version) %>%
    semver::parse_version() %>%
    as.data.frame() 
    ) %>%
  arrange(major, minor, patch)
```

The cleaned dataset contains the following 12 columns: package_name, version, time_stamp, date, year, month, weekday, major, minor, patch, prerelease, and build.

## Analysis

### Descriptive statistics (all versions)

The [dependency manifest file](https://github.com/canada-ca/tracker/blob/003a3d6e0452fb9a248a7d51fd5556bc03dfbcc6/frontend/package.json) uses a total of 72 packages, which, all together, were updated through major, minor, patches and pre-releases in a total of 7,322 versions since December 29th 2010. Of those, 2,660 (36%) were pre-releases which were removed as they are not expected to significantly impact development and maintenance of the [dependency manifest file](https://github.com/canada-ca/tracker/blob/003a3d6e0452fb9a248a7d51fd5556bc03dfbcc6/frontend/package.json). 

```{r, include=FALSE}
data <- data %>% 
  filter(prerelease == "") 
```

Figure 1 is a plot of the number of versions for each package in this time period, with the package with the most versions at the top (webpack had 307 versions) and the package with the least versions at the bottom (webpack-config-utils had six versions). Figure 1 illustrates the variation in the number of versions among packages. Note that this might be due to the creation date of packages as they did not all exist in December 2010.

```{r, fig.height=18, echo=FALSE}
data %>% 
  count(package_name, sort = TRUE) %>%
  mutate(package_name = fct_reorder(package_name, n)) %>%
  ggplot(aes(n, package_name)) +
  geom_col() +
  labs(
    title = "Figure 1: Number of versions for each package",
    x = "Number of versions (majors, minors, patches and pre-releases)", 
    y = "Package name"
  ) + 
  theme_minimal()
```

Figure 2 is a heatmap of the number of versions per year per package with packages organised in alphabetical order on the y-axis. The more versions in a year for a given package, the darker the color. This figure illustrates that packages have been created in every year between 2011 and 2019 and that the number of versions per package per year is highly variable. Some packages had most of their versions in the early years of their creation (e.g., babel-core), others had most versions in recent years (e.g., react) and others had no clear trends (e.g., d3). Note that at the time of writing this, the year 2020 was not complete.  

```{r, fig.height=18, echo=FALSE}
heatmap_data <- data %>% group_by(year, package_name) %>% tally()

heatmap <- heatmap_data %>% 
  ggplot(mapping = aes(x = year, y = reorder(package_name, desc(package_name)), fill = n)) +
  geom_tile() +
  labs(
    title = "Figure 2: Number of versions per year per package",
    x = "Year", 
    y = "Package name"
  ) +
  scale_fill_gradient(name = "# of versions",
                      low = "#e8f1fa",
                      high = "#002e5c") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))

heatmap
```

Figure 3 illustrates the total number of versions per year between December 2010 and November 2020 for the packages used in the [dependency manifest file](https://github.com/canada-ca/tracker/blob/003a3d6e0452fb9a248a7d51fd5556bc03dfbcc6/frontend/package.json). Note that at the time of writing this, the year 2020 was not complete. Between 2011 and 2019, there was an increasing number of versions per year, partially due to new packages being created.

```{r, echo=FALSE}
data %>%
  group_by(year) %>%
  tally() %>%
  ggplot(aes(year, n)) +
  geom_col() +
  theme_classic() +
  labs(
    title = "Figure 3: Number of versions per year",
    x = "Year", 
    y = "Number of versions (major, minor and patches)"
  )
```

Figure 4 illustrates the number of packages used in the [dependency manifest file](https://github.com/canada-ca/tracker/blob/003a3d6e0452fb9a248a7d51fd5556bc03dfbcc6/frontend/package.json) that existed per year between December 2010 and November 2020. Note that at the time of writing this, the year 2020 was not complete. Between 2011 and 2019, there was an increasing number of packages for which versions were being deployer per year.

```{r, echo=FALSE}
data %>%
  group_by(year) %>%
  summarise(number_of_packages = n_distinct(package_name)) %>%
  ggplot(aes(year, number_of_packages)) +
  geom_col() +
  theme_classic() +
  labs(
    title = "Figure 4: Number of packages per year",
    x = "Year", 
    y = "Number of packages"
  )
```

Figure 5 illustrates the number of versions per package used in the [dependency manifest file](https://github.com/canada-ca/tracker/blob/003a3d6e0452fb9a248a7d51fd5556bc03dfbcc6/frontend/package.json) that existed per year between December 2010 and November 2020. Note that at the time of writing this, the year 2020 was not complete.

```{r, echo=FALSE}
data %>%
  group_by(year) %>%
  summarise(
    number_of_versions = n(),
    number_of_packages = n_distinct(package_name),
    number_versions_per_package = number_of_versions / number_of_packages
            ) %>%
  ggplot(aes(year, number_versions_per_package)) +
  geom_col() +
  theme_classic() +
  labs(
    title = "Figure 5: Number of versions per package per year",
    x = "Year", 
    y = "Number of versions per package"
  )
```

### Time between major versions

In this section, we are interested in the time between major versions. In theory, minor versions and a patches will not require code changes but major versions will. We therefore only selected the major versions from the datasets. There was a total of 262 major versions, representing less than 4% of all versions.

```{r, include=FALSE}
major_versions_only <- data %>% 
  filter(minor == "0") %>% 
  filter(patch == "0")
```

```{r, include=FALSE}
interval_between_major_versions_only <- major_versions_only %>%
   mutate(major_version = paste0(package_name, " ", major, ".", minor, ".", patch)) %>% 
   arrange(time_stamp) %>%
   mutate(
     days_until_next_major_version = as.numeric(lead(date) - date)
   ) %>%
  drop_na(days_until_next_major_version) 
```


```{r, echo=FALSE}
interval_between_major_versions_only %>% 
  ggplot(
    aes(x=days_until_next_major_version)
    ) +
    geom_histogram() + 
  theme_minimal() +
  labs(
    title = "Figure 6: Distribution of time between major versions (all years)",
    x = "Time (days) between major versions",
    y = "Count"
  )
```

```{r, echo=FALSE}
interval_between_major_versions_only %>% 
  ggplot(
    aes(x=days_until_next_major_version)
    ) +
    geom_histogram() + 
    theme_minimal() +
  labs(
    title = "Figure 7: Distribution of time between major versions (by year)",
    x = "Time (days) between major versions",
    y = "Count"
  ) +
  facet_wrap(~ year)
```

```{r, include=FALSE}
table <- interval_between_major_versions_only %>% 
  group_by(year) %>%
  summarise(
    number_of_major_versions = n(), 
    median_time = median(days_until_next_major_version),
    average_time = mean(days_until_next_major_version),
    standard_deviation = sd(days_until_next_major_version)
  ) 
```

```{r, echo=FALSE}
formattable(
  table,
  align =c("l","c","c","c","c")
  )
```

Figure 7 demonstrates the differences between The distributions (Figure 7, the number of versions per year and the average time between major versions of days between major versions changeaverage time between major versions 

```{r, echo=FALSE}
interval_between_major_versions_only %>% 
  filter(date >= "2018-01-01") %>%
  filter(date < "2020-01-01") %>%
  ggplot(
    aes(x=days_until_next_major_version)
    ) +
    geom_histogram() + 
    theme_minimal() +
  labs(
    title = "Figure 8: Distribution of time between major versions (2018 and 2019)",
    x = "Time (days) between major versions",
    y = "Count"
  ) +
  facet_wrap(~ year)
```

```{r, echo=FALSE}
interval_between_major_versions_only %>% 
  filter(date >= "2018-01-01") %>%
  filter(date < "2020-01-01") %>%
  ggplot(
    aes(x=days_until_next_major_version)
    ) +
    geom_histogram() + 
    theme_minimal() +
  labs(
    title = "Figure 9: Distribution of time between major versions (2018 and 2019)",
    x = "Time (days) between major versions",
    y = "Count"
  ) 
```

```{r}
interval_between_major_versions_only %>% 
  filter(date >= "2018-01-01") %>%
  filter(date < "2020-01-01") %>%
  summarise(
    number_of_major_versions = n(), 
    median_time = median(days_until_next_major_version),
    average_time = mean(days_until_next_major_version),
    standard_deviation = sd(days_until_next_major_version)
  ) 
```