---
title: "Software dependancy analysis"
author: "C. Mimeault and M. Williamson"
date: "11/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

```{r libraries, include=FALSE}
library(tidyverse)
library(lubridate)
library(semver)
library(scales)
library(skimr)
```

## Context

Open source projects rely on several libraries that get updated regularly. While not all updates have impact on the code that relies on it, some do. 

According to semantic versioning (see <https://semver.org>), versions are numbered incrementally according to the type of change. A "MAJOR version" signifies incompatible API changes, a "MINOR version" signifies added functionality in a backwards compatible manner, and a "PATCH version" signifies backwards compatible bug fixes.

This project explores the number and frequency of updates in a project currently under development. It aims to shed light on the update requirements once the project will be in production.

## Methods

mon amour - can you write the section on how you extracted the data?

On October 22, 2020, the registry for NPMJS...

`
curl -sL "https://registry.npmjs.org/$pkg" | jq -r --arg  name "$pkg" '.time | to_entries[] | [$name, .key, .value] | @csv'
`

The returned data was saved as a csv file. It contained three columns containing the name of the package, the version and a time stamp.

## Cleaning data

The data needed a little attention before any analysis would be possible. The first step was to read the csv file including names for each column as those were not specififed in the dataset. 
```{r data, message=FALSE}
packages <- read_csv("~/projects/dependency-analysis/data/raw/nodejs.csv", col_names = c("package_name", "version", "time_stamp"))
```

Then, we removed rows that did not contain data, extracted the date from the time stamp, and parsed the version number to only include the first three numbers and the dots separating the numbers.

```{r}
data <- packages %>%
  filter(version != "created") %>%
  filter(version != "modified") %>%
  mutate(
    date = as.Date(time_stamp, "%Y/%m/%d"),
    year = as.factor(year(date)),
    month = as.factor(month(date)),
    weekday = as.factor(weekdays(date)),
    version = str_replace(version, "^(\\d{1,2}.\\d{1,2}.\\d{1,2})(\\w+)$", "\\1-\\2")
  )
```

Finally, using the semver library, we extracted the version numbers into separate columns for major, minor and patch versions.

```{r}
data <- add_column(
  data,
  pull(
    data, version) %>%
    semver::parse_version() %>%
    as.data.frame() 
    ) %>%
  arrange(major, minor, patch)
```

This left us with a dataset with nine columns being:
```{r}
colnames(data)
```

## Analysis

###  Number of versions per date
```{r}
# plot all versions per date
data %>%
  group_by(date) %>%
  tally() %>%
  ggplot(aes(date, n)) +
  geom_line() +
  theme_classic() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    x = "Date", 
    y = "Number of versions (major, minor and patches)"
  )
```

### Number of versions per year
```{r}
# plot all versions per date
data %>%
  group_by(year) %>%
  tally() %>%
  ggplot(aes(year, n)) +
  geom_col() +
  theme_classic() +
  labs(
    x = "Year", 
    y = "Number of versions (major, minor and patches)"
  )
```

### Number of versions per year
```{r}
# plot all versions per date
data %>%
  group_by(month) %>%
  tally() %>%
  ggplot(aes(month, n)) +
  geom_col() +
  theme_classic() +
  labs(
    x = "Month", 
    y = "Number of versions (major, minor and patches)"
  )
```


```{r}
# plot all versions per date
data %>%
  group_by(weekday) %>%
  tally() %>%
  ggplot(aes(weekday, n)) +
  geom_col() +
  theme_classic() +
  labs(
    x = "Weekday", 
    y = "Number of versions (major, minor and patches)"
  )
```

```{r}
library(reshape)
table <- cast(data, package_name ~ year)
table
```


```{r}
library(ggplot2)

heatmap_data <- data %>% group_by(year, package_name) %>% tally()

heatmap <- heatmap_data %>% 
  ggplot(mapping = aes(x = year, y = package_name, fill = n)) +
  geom_tile() +
  xlab(label = "Year") +
  scale_fill_gradient(name = "Number of versions",
                      low = "#FFFFFF",
                      high = "#012345") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

heatmap
```


